\documentclass[a4paper,12pt,twoside,openright]{memoir}

\input{Required/specifications.tex}
\input{Required/abbreviations.tex}

\begin{document}

% \input{help.tex} % Comment this to remove the help pages

%------------------------------------------------------
%   ↓   INTRODUCTION    ↓
%------------------------------------------------------

\chapter[Randomizing Language Features][Tamgwa]{Randomizing Language Features in Tamgwa}

\byline{armytag} % add your name as you wish to be credited
\articlesub{When and how to roll the dice} % use this command to insert a subtitle

\thispagestyle{fancy}
\BgUsetrue

%------------------------------------------------------
%   ↓   ARTICLE BODY    ↓
%------------------------------------------------------

    For a long time I have struggled to get very far with making a conlang, too bewildered by the vast sea of linguistic possibilities to actually choose how this one language in front of me should work.  So for my newest language \textbf{Tamgwa}, I decided to take the process out of my head until it's take enough shape for me to work with.  In other words, I randomly selected a lot of features of \textbf{Tamgwa} and started to fill in the gaps and smooth out the edges based on what was generated.  There was no Artificial Intelligence involved in this process, just good old fashioned chance which lead to some interesting developments along the way.  I hope this article will give you some ideas about how to use randomization in your own language creation process.

\section*{Why use randomization?} % * for unnumbered section

    At first glance, randomization might seem at odds with the very spirit of conlanging.  Whether we make languages to flesh out our fictional worlds, to explore the edges of linguistic and philosophical expression, to further the noble cause of international communication, or simply for the glee this process brings us, we never want what we do to be a random mess.  There is almost always a \textit{purpose} behind our endeavor, an idea pushing forward all the effort we inevitably dedicate to this pursuit.  If all we cared about were the phonologies, inflection tables, and word lists, we might spend a few hours with Vulgarlang or ChatGPT seeing what combinations they spit out.\footnote{If this actually is what you're interested in, I recommend looking at real languages anyway.  The diversity in natural languages far outshines anything these tools can offer.}  But that kind of activity is never going to sustain our interest for very long, certainly not for the months and years that many conlangers put into their creations.  Clearly there is something to this process fuels the passion of a conlanger, which randomization along can never fulfill.

    So is there any place for randomization in the conlanging process?  The answer, of course, is yes!  Randomization works best as a tool in a conlanger's toolbelt, augmenting their creation process without dominating it.  This can happen in several ways.  First, randomization can help a conlanger generate content like phonemes, words, grammar structures, and more.  This benefit should be obvious to anyone who winces remembering the half-finished dictionary they haven't touched in months. Second, randomization can counter a conlanger's linguistic biases.  Few of us are familiar with more than a handful of languages, and that limited exposure can likewise limit the form our conlangs take.  If you can't help but make Englishy conlangs, chance can steer you elsewhere.  Lastly, randomization can help overcome the indecision we sometimes feel in the face of endless linguistic possibility.  There are so many ways that human language conveys ideas, and sometimes it can be almost impossible to wade through the myriad options to settle on the \textbf{one} this language will use.

    All these situations and more are excellent times to use a bit of randomization.  Just remember that randomization is only a tool to assist with the actual goal: creating a language.

\section*{Laying the foundation} % * for unnumbered section

    Now that I've convinced you of the merits of a bit of randomization, you still need to know how to use it.  To that end, I will walk through the process I used for starting \textbf{Tamgwa} with almost entirely randomized features.  The simplest way to randomize a feature is to list a few variations of how that feature could look, and then toss a coin, roll some dice, or otherwise randomly choose which variation to move forward with.  For example, we could take the list of the types of writing systems from Wikipedia (abjad, abugida, alphabetical, and logographic/syllabic) and roll a 4-sided die to decide which system to implement.  The result might be at odds with the structure or phonology of the language, such as a syllabary for language with complex syllable structure, but that is precisely how randomization challenges our biases in language construction.  We can always put the random result aside if it simply isn't working with the rest of the language.

    For \textbf{Tamgwa}, rather than using Wikipedia I instead used the invaluable World Atlas of Langauge Structures (WALS).\footnote{https://wals.info}  WALS has a collection of chapters and associated maps showing the distribution of a variety of language features, from phonology to morphology to syntax and more.  Each chapter lists various realizations of the language feature along with the number of languages using that variation.  These numbers give an idea of how widespread each variation is and allow us to weight the randomization in a more naturalistic way.  In other words, we can use the chapters in WALS to randomize an array of language features in a way that mimics the tendencies of natural languages.

    To illustrate this process, we'll randomize how \textbf{Tamgwa} encodes noun plurality using WALS chapters 33 \& 34.  Chapter 33 covers how plurality is encoded, such as using affixes like English's \textit{-s}.  The authors list the following possibilities and number of languages which employ that option.

\begin{table}[H]
	\centering
	% \begin{tabu}{$>{\bfseries}l|^c^c^c^c^c}
	\begin{tabu}{>{\bfseries}l|^c^c}
	\rowstyle{\bfseries}
        Language Feature & Language Count & Randomizer Range \\
		\hline
        Plural prefix & 126 & 1 - 126 \\
        Plural suffix & 513 & 127 - 639 \\
        Plural stem change & 6 & 640 - 645 \\
        Plural tone & 4 & 646 - 649 \\
        Plural by complete stem reduplication & 8 & 650 - 657 \\
        Morphological, but no method primary & 60 & 658 - 717 \\
        Plural word & 170 & 718 - 887 \\
        Plural clitic & 81 & 888 - 968 \\
        No plural & 98 & 969 - 1066\\
	\end{tabu}
	\caption{WALS Chapter 33: Coding of Nominal Plurality}
	\label{wals-33}
\end{table}

    I have added the Randomizer Range column, which simply shows how our random number should map back to the feature options.\footnote{To create the Randomizer Ranges yourself, simply take the high end of the previous range, add 1 to it to get the low end of the new range, and add the number of languages to it to get the high end of the new range}  Now all we have to do is generate a number between 1 and 1066 (inclusive) and see which range it falls within.  For \textbf{Tamgwa} I generated the number 814, so \textbf{Tamgwa} will encode plurality with an independant word rather than some kind of morphological inflection.  We can use the same process on Chapter 34: Occurrence of Nominal Plurality to randomize \textit{when} this plural word should be used.  Hopefully we don't randomly select the option "No nominal plural", since we've already established that \textbf{Tamgwa} will use a plural word; we could either not include this option or re-randomize if we get it.  Cutting to the end, I ultimately generated the random number 118 for this feature which corresponds to the option "Plural in all nouns, always optional".  Thus in \textbf{Tamgwa} all nouns can be optionally marked for plurality by adding a plural word somewhere in the sentence.

    For \textbf{Tamgwa} I used this process to randomize features for most of the WALS chapters.  This provides a wide grammatical basis upon which to build the rest of the language.  I've listed some of the randomized features below.  Of course some contradictions are likely to arise from these completely random features, which can be seen in this list as well.  These contradictions are not always obstacles however, and they can sometimes lead to developing interesting new features to resolve them.  I will go into an example of this next.

\begin{itemize}
    \item 1
    \item 2
\end{itemize}

\section*{Personal disagreement} % * for unnumbered section

    The randomized features of Tamgwa are (surprisingly) mostly consistent, outlining an analytic language with very little inflectional morphology.  There are some still some inconsistencies however, the most glaring of which is between the features for chapters 26 and 102.  Verbs in Tamgwa must have "person marking of both A and P arguments" according to the chapter 102 feature.  Yet in order for Tamgwa to have "little to no inflectional morphology", the \textbf{affix index} must be at most 2.  According to the description in chapter 26, affixes marking A add 2 to the affix index and affixes marking P add an additional 1, so it would seem that Tamgwa must necessarily have an affixing index of at least 3 which would fully contradict the value generated for this feature.

    Complete contradictions like this could be easily overcome by simply rerandomizing some or all of the features.  In this case however, there are some details in the WALS definitions that can help resolve the contradiction.  In chapter 26, the authors clarify that person-marking clitics that can also attach to non-verbs are not included as part of the affix index calculation.  On the other hand, chapter 102 does count clitics in its analysis as long as the clitics can sometimes attach to the verb.  Therefore, if at least one of the person markers is a clitic that sometimes attaches to verbs and sometimes to non-verbs, then the affix index can be kept to 2 or less and both of these features can be satisfied.

    An real-world example of a language with verbal person marking that uses affixes and clitics together is the Sorani dialect of Kurdish.  According to the analysis by Gharib and Pye\footnote{Gharib, H., \& Pye, C. (2023) The clitic status of person markers in Sorani Kurdish. Kansas Working Papers in Linguistics, 39, 57-65. https://doi.org/10.17161/1808.27692}, one of the verb arguments is marked by an affix on the verb and the other person is marked by a clitic on either the verb or a preceding adverb.  Whether the clitic is marking the agent (A) or the patient (P) of the action depends on whether it is using the Perfective or Imperfective aspect, as seen in the example below.

\begin{examples}
    \ex
    \lect Sorani Kurdish
    \words \MC2 dat \MC2 benim
    \bits da =t beni -m
    \gloss IPFV OBL.2SG see.IPFV 1SG
    \tr I see you.
    \ex
    \words \MC3 benemeet
    \bits beni =m -eet
    \gloss see.PFV OBL.1SG 2SG
    \tr I saw you.
    \ex
    \words \MC3 benetm
    \bits bene =t -m
    \gloss see.PFV OBL.2SG 1SG
    \tr You saw me.
    \source Gharib \& Pye 2023
\end{examples}

    In Sorani Kurdish the oblique clitic is used to mark the patient of imperfective verbs and the agent of perfective verbs.  \textbf{Tamgwa} uses a simpler system, where the clitic attaches to the first verb/adverb in the verb phrase regardless of aspect.  Thus for our purposes, it is sufficient to see that examples (1) and (3) show the OBL.2SG clitic attaching to the aspect marker if it is present, and the verb if not.

\section*{Producing Pronouns}

    Now that we have an idea of how verbal person marking will work in \textbf{Tamgwa}, we're ready to pick some pronouns and put the system to the test.  Of course this process will involve some randomization too!  There are plenty of ways to generate a list of words and syllables, such as the online tool Vulgarlang\footnote{https://www.vulgarlang.com}, but I've written a simple Python script that will give me a certain number of words based on the phonology and phonotactics of \textbf{Tamgwa}.

    Based on the features we generated at the beginning, \textbf{Tamgwa} will need indepent pronoun words with person-number stems including a clusivity distinction in 1PL pronouns.  Since we will be marking agent and patient on the verb, we may as well make separate nominative and accusative forms of the pronouns as well.  Finally, we will also indicate possession with a specially inflected pronoun, so we also need some kind of genetive form.  Altogether that makes 7 person-number combinations and 5 forms (3 cases, 1 affix, and 1 clitic form) for a total of 35 distinct "words".  After telling the Python script to generate 35 words we can put the results into a table to better visualize the pronouns we've just created.

\begin{table}[H]
	\centering
	% \begin{tabu}{$>{\bfseries}l|^c^c^c^c^c}
	\begin{tabu}{>{\bfseries}l|^c^c^c^c^c}
	\rowstyle{\bfseries}
        & NOM & Affix & ACC & Clitic & GEN \\
		\hline
        1SG     & hnja      & ki    & gya       & hnja  & hma    \\
        2SG     & lam       & man   & ba        & be    & pu     \\
        3SG     & le        & na    & bwa       & nji   & pe     \\
        1PL.INC & ga        & fe    & hmuhl     & pix   & re     \\
        1PL.EXC & pwa       & tenj  & hla       & le    & pu     \\
        2PL     & bihl      & tu    & bya       & bes   & hnjanj \\
        3PL     & xa        & ti    & pwax      & njar  & bwa    \\
	\end{tabu}
	\caption{Random pronoun table}
	\label{rand-pronoun}
\end{table}

    This table is, obviously, totally random and nonsensical.  Keep in mind, though, that the randomization is merely meant to give us a starting point.  From here we can look for patterns and rearrange the table to fit it into something more reasonable.  Personally, I immediately latched onto the two \textbf{hnja} in the 1SG row and decided to move all the words beginning with \textbf{hnj} or \textbf{nj} into that row.  From there we continue forming other patters, such as voiceless labial onsets for 2SG and closed syllables for GEN.  The arrangement I finally settled on is printed below, with obligatory plural marker \textbf{tyan}\footnote{\textit{N.B.}\textbf{tyan} is an optional plural marker for all nouns, but is only obligatory for plural pronouns.  The feature was randomly generated for Chapter 35 of WALS.} added to NOM and ACC forms of plural pronouns.

\begin{table}[H]
	\centering
	% \begin{tabu}{$>{\bfseries}l|^c^c^c^c^c}
	\begin{tabu}{>{\bfseries}l|^c^c^c^c^c}
	\rowstyle{\bfseries}
        & NOM & Affix & ACC & Clitic & GEN \\
		\hline
        1SG     & hnja      & -nji  & hnja      & =njar & hnjanj \\
        2SG     & fe        & -pe   & pwax      & =pwa  & pix    \\
        3SG     & bes       & -hla  & hmuhl     & =hma  & man    \\
        1PL.INC & tu tyan   & -ti   & re tyan   & =na   & tenj   \\
        1PL.EXC & xa tyan   & -ki   & gya tyan  & =ga   & bya    \\
        2PL     & bwa tyan  & -be   & bwa tyan  & =ba   & bihl   \\
        3PL     & pu tyan   & -le   & pu tyan   & =le   & lam    \\
	\end{tabu}
	\caption{Rearranged pronoun table}
	\label{arr-pronoun}
\end{table}

\section*{Revealing Results}

    Now that we have a full set of personal pronouns generated, we just need some preverbal tense-aspects words for the clitics to attach to, and then we'll be able to fully test our new syntax.  I've decided to have \textbf{Tamgwa} use the same set of aspect words as described in Timothy Ajani's 2001 paper.\footnote{Ajani, T. 2001. \textit{Aspect in Yoruba and Nigerian English}}  According to this analysis, we will need 4 aspect words marking Realis Imperfective, Irrealis Perfective, and Irrealis Imperfective, and Relational aspects.  The Realis Perfective is unmarked and thus assumed when none of these explicit aspect words occur.  The aspect words can co-occur to create more nuanced aspectual distinctions, such as Realis Imperfective and Irrealis Imperfective together denoting a Habitual aspect.  For more details on how this works, please see Ajani's analysis on the original Yoruba system.

    Unlike when we generated the whole set of personal pronouns and reorganized them afterward, when I'm creating just a handful of words I like to generate 2-3 times the number of words I need and manually pick from the resulting list.  The results are listed in the table below.  

\begin{table}[H]
	\centering
	% \begin{tabu}{$>{\bfseries}l|^c^c^c^c^c}
	\begin{tabu}{>{\bfseries}l|^c}
	\rowstyle{\bfseries}
        Aspect & Word \\
		\hline
        RLS.PFV  & <none> \\
        RLS.IPFV & kwa    \\
        IRR.PFV  & hna    \\
        IRR.IPFV & pi     \\
        REL      & gya    \\
	\end{tabu}
	\caption{Aspect words in Tamgwa}
	\label{aspect-words}
\end{table}

    With both the personal pronouns and aspect words, we can finally test how the person-marking clitic interacts with the verbal aspect system.  We've decided that the clitics will attach at the end of the first word in the aspect-verb structure.  Below are some examples of what that looks like.

\begin{examples}
    \ex
    \lect Tamgwa
    \words \MC3 kyamunjipwa
    \bits kyamu -nji =pwa
    \gloss see 1SG.NOM 2SG.ACC
    \tr I saw you.
    \ex
    \words \MC2 kwapwa \MC2 kyamunji
    \bits kwa =pwa kyamu -nji
    \gloss RLS.IPFV 2SG.ACC see 1SG.NOM
    \tr I see you.
    \ex
    \words \MC2 pipwa \MC2 kyamunji
    \bits pi =pwa kyamu -nji
    \gloss IRR.IPFV 2SG.ACC see 1SG.NOM
    \tr I will see you.
    \ex
    \words \MC2 pipwa kwa \MC2 kyamunji
    \bits  pi =pwa kwa kyamu -nji
    \gloss IRR.IPFV 2SG.ACC RLS.IPFV see 1SG.NOM
    \tr I generally (habitually) see you.
\end{examples}

\section*{Conclusion} % * for unnumbered section

    Hopefully this article has given you some ideas about how to use randomization to augment your conlanging process.  You can randomize everything from syntax to morphology to phonology to lexicon and more.  This is particularly useful if you're struggling with choice paralysis or not sure where to start with certain part of your language, and you can even go as far as bootstrapping most of a language's features through random generation.  You can also go about doing your randomization in whatever way you'd like.  You write down some options on flash cards to randomly draw from.  However you choose to use randomization, if at all, I wish you luck and can't wait to see what you come up with.

\end{document}
